Wait-free Queues with Polylogarithmic Step Complexity

### 2022-09-30 thesis/thesis.tex
add lemma sum enq

\paragraph{}
Starting from the root, \textsc{GetIndex}($i$) BinarySearches $i$ in the prefix sum array to find block containing $i$th operation, then continues recursively calling \textsc{GetElement}($b,i$) to find $i$th element of block $b$. From lemma $\ref{lem:block_size}$ we know a block size is at most $p$. So BinarySearch takes at most \textsc(O)$(\log p)$, since  with knowing pointers of a block and its previous block we can determine the base \texttt{(domain ?)} to search and its size is \textsc{O}$(p)$.

@@

\paragraph{}
In the Block Tree, we did not store the sequence of operations explicitly but instead stored blocks of concurrent operations to optimize \texttt{Propagate()} steps and increase parallelism. So now the problem is to find the result of each Dequeue. From lemma \ref{lem:block_size} we know we can linearize operations in a block in any order; here, we choose to decide to put Enqueue operations in a block before Dequeue operations. In the next example, operations in a cell are concurrent. \texttt{DEQ()} operations return \texttt{null, 5, 2, 1, 3, 4, null} respectively. We will next describe how these values can be computed efficiently.

@@

\begin{definition}[Linearization] \label{def::lin}
 $$L=E(root,1)\cdot D(root,1)\cdot E(root,2)\cdot D(root,2)\cdot E(root,3)\cdot D(root,3)\cdots$$
\end{definition}
\pagebreak

@@

Now we prove some claims about the size and operations of a block. These lemmas will be used for analysis and correctness of \nf{GetEnqueue()}.

@@

For the induction step we prove if \nf{$n$.child.GetEnqueue($sb$, i)} returns $E_i(n.child, sb)$ then \nf{$n$.GetEnqueue($b$, $i$)} returns $E_i(n,b)$.  From Definition \ref{ordering} of $E(n,b)$, operations from the left subblocks come before the operations from the right subblocks in a block (see Figure \ref{figGet}). \nf{num\sub{enq-left}} field  in \nf{$n$.blocks[$b$]} is the number of \nf{Enqueue()} operations from the blocks's subblocks in the left child of \nf{n}. So the $i$th enqueue operation is propagated from the right child if and only if $i$ is greater than \nf{$b$.num\sub{enq-left}}. Line \ref{leftOrRight} decides whether the \nf{i}th enqueue in  $b$th block  of internal node \nf{n} is in the  left child or right child subblocks of \nf{n.blocks[b]}. By Definitions \ref{def::ops}, \ref{def::subblock} to find an operation in subblocks of \nf{$n$.blocks[$i$]} we need to search in the range
\begin{align*}
&\texttt{$n$.left.blocks[$n$.blocks[$i$-1].end\textsubscript{left}+1..$n$.blocks[$i$].end\textsubscript{left}]}  \cup \\
&\texttt{$n$.right.blocks[$n$.blocks[$i$-1].end\textsubscript{right}+1..$n$.blocks[$i$].end\textsubscript{right}]}.
 \end{align*}
First we consider the case where the \nf{Enqueue} we are looking is in the left child. There are $eb=n.\nf{blocks[}b-1\nf{].sum\sub{enq-left}}$ enqueues in the blocks before the left subblocks of $n$\nf{.blocks[$b$]}, so $E_i(n,b)$ is $E_{i+eb}(n.left)$ which is $E_{i^\prime}(n.left,b^\prime)$ for some $b^\prime$ and $i^\prime$. We can compute $b^\prime$ and then search for the $i+eb$th enqueue in \tt{n.left}, where $i^\prime$ is \tt{i+eb-n.left.blocks[}$b^\prime-1$\tt{].sum\sub{enq}}. The parameters in Line \ref{leftChildGet} are for searching $E_{i+eb}(n.left)$ in \nf{n.left.block} in the range of left subblocks of $n$\nf{.blocks[}$b$\nf{]}, so this \nf{BinarySearch} returns the index of the subblock containing $E_i(n,b)$.

@@

\begin{lemma}
An \nf{Enqueue()} or \nf{Dequeue()} operation, does at most $4\log p$ \nf{CAS} operations.
\end{lemma}
\begin{proof}
  In each height of the tree at most 2 times \nf{Refresh()} is invoked and  every \nf{Refresh()} has 2 \nf{CAS}es, one in Line \ref{cas} and one in Lines \ref{incrementHead1} or \ref{incrementHead2}.
\end{proof}

>>>>

\paragraph{}
Starting from the root, \textsc{GetIndex}($i$) BinarySearches $i$ in the prefix sum array to find block containing $i$th operation, then continues recursively calling \textsc{GetElement}($b,i$) to find $i$th element of block $b$. From lemma $\ref{blockSize}$ we know a block size is at most $p$. So BinarySearch takes at most \textsc(O)$(\log p)$, since  with knowing pointers of a block and its previous block we can determine the base \texttt{(domain ?)} to search and its size is \textsc{O}$(p)$.

@@

\paragraph{}
In the Block Tree, we did not store the sequence of operations explicitly but instead stored blocks of concurrent operations to optimize \texttt{Propagate()} steps and increase parallelism. So now the problem is to find the result of each Dequeue. From lemma \ref{blockSize} we know we can linearize operations in a block in any order; here, we choose to decide to put Enqueue operations in a block before Dequeue operations. In the next example, operations in a cell are concurrent. \texttt{DEQ()} operations return \texttt{null, 5, 2, 1, 3, 4, null} respectively. We will next describe how these values can be computed efficiently.

@@

\begin{definition}[Linearization] \label{def::lin}
 $$L=E(root,1)\cdot D(root,1)\cdot E(root,2)\cdot D(root,2)\cdot E(root,3)\cdot D(root,3)\cdots$$
\end{definition}

The next two observations are correct for both enqueues and deqeueues.

\begin{observation}
  $n\nf{.blocks[}b\nf{]}\nf{.num\sub{enq}}$ is the size of $E(n,b)$.
\end{observation}

\begin{observation}\label{lem::sumDeq}
$\sum_{i=1}^{b} {E(n,i)}=n\nf{.blocks[}i\nf{].sum\sub{enq}}$.
\end{observation}

The next observation is correct for both directions.

\begin{observation}\label{lem::numEnqLeft}
Let $B$ be $n\nf{.blocks[}b\nf{]}$. $B\nf{.num\sub{enq-left}}$ is the number of \nf{Enqueue} operations in the direct subblocks of $B$ in $n$\nf{.left}.
\end{observation}
\begin{proof}
Let $B^\prime$ be $n\nf{.blocks[}b-1\nf{]}$. We prove the claim by induction on height of node $n$. 
\begin{align*}
    B\nf{.num\sub{enq-left}}=&B.\nf{sum\sub{enq-left}}-B^\prime.\nf{sum\sub{enq-left}}& \text{Definition of }\nf{num\sub{x}}\\
    =&B^\prime.\nf{sum\sub{enq-left}}+n\nf{.left.blocks[}B\nf{.end\sub{left}].sum\sub{enq}}\\&-n\nf{.left.blocks[}B^\prime\nf{.end\sub{left}].sum\sub{enq}}-B^\prime.\nf{sum\sub{enq-left}} & \nf{CreateBlock} \text{ code}\\
    =&\sum_{i=B^\prime\nf{.end\sub{left}}+1}^{B\nf{.end\sub{left}}}  n\nf{.left.blocks[}i\nf{].num\sub{enq}}
\end{align*}
\end{proof}
\pagebreak

@@

First we prove some claims about the size and operations of a block. These lemmas will be used later for correctness and analysis of \nf{GetEnqueue()}.

@@

For the induction step we prove if \nf{$n$.child.GetEnqueue($sb$, i)} returns $E_i(n.child, sb)$ then \nf{$n$.GetEnqueue($b$, $i$)} returns $E_i(n,b)$.  From Definition \ref{ordering} of $E(n,b)$, operations from the left subblocks come before the operations from the right subblocks in a block (see Figure \ref{figGet}). By Observation \ref{lem::numEnqLeft}, the \nf{num\sub{enq-left}} field  in \nf{$n$.blocks[$b$]} is the number of \nf{Enqueue()} operations from the blocks's subblocks in the left child of \nf{n}. So the $i$th enqueue operation is propagated from the right child if and only if $i$ is greater than \nf{$b$.num\sub{enq-left}}. Line \ref{leftOrRight} decides whether the \nf{i}th enqueue in  $b$th block  of internal node \nf{n} is in the  left child or right child subblocks of \nf{n.blocks[b]}. By Definitions \ref{def::ops}, \ref{def::subblock} to find an operation in subblocks of \nf{$n$.blocks[$i$]} we need to search in the range
\begin{align*}
&\texttt{$n$.left.blocks[$n$.blocks[$i$-1].end\textsubscript{left}+1..$n$.blocks[$i$].end\textsubscript{left}]}  \cup \\
&\texttt{$n$.right.blocks[$n$.blocks[$i$-1].end\textsubscript{right}+1..$n$.blocks[$i$].end\textsubscript{right}]}.
 \end{align*}
First we consider the case where the \nf{Enqueue} we are looking is in the left child. There are $eb=n.\nf{blocks[}b-1\nf{].sum\sub{enq-left}}$ enqueues in the blocks before the left subblocks of $n$\nf{.blocks[$b$]}, so $E_i(n,b)$ is $E_{i+eb}(n.left)$ which is $E_{i^\prime}(n.left,b^\prime)$ for some $b^\prime$ and $i^\prime$. We can compute $b^\prime$ and then search for the $i+eb$th enqueue in \tt{n.left}, where $i^\prime$ is \tt{i+eb-n.left.blocks[}$b^\prime-1$\tt{].sum\sub{enq}}. The parameters in Line \ref{leftChildGet} are for searching $E_{i+eb}(n.left)$ in \nf{n.left.block} in the range of left subblocks of $n$\nf{.blocks[}$b$\nf{]}, so this \nf{BinarySearch} returns the index of the subblock containing $E_i(n,b)$.

@@

\begin{lemma}
An \nf{Enqueue()} or \nf{Dequeue()} operation, does at most $4\log p$ \nf{CAS} operations.
\end{lemma}
\begin{proof}
  In each height of the tree at most 2 times \nf{Refresh()} is invoked and  every \nf{Refresh()} invokes at most 3 \nf{CAS}es, one in Line \ref{cas} and two from \nf{Advance} in Line \ref{incrementHead}.
\end{proof}
